{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-06-01 ms.tex
Add graphicx in
Somehow this is not added?  I thought it was and it doesn't compile
here without.  But then I don't see a change in the log.  Issue
is possibly due to a minimal tex installation here.
Spelling how does it work
Typo-squashing edits throughout.
Mostly from Andrea's read-through

\newcommand{\plant}{\texttt{plant}}

\usepackage[colorinlistoftodos]{todonotes}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how  
  physiological rules and functional trade-offs interact with competition and 
  disturbance regimes to influence vegetation demography, structure and 
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and 
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; to maintaining support tissues \citep{Givnish-1988,
  Enquist-2007} and to reproduction \citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
Ontogenetic 
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates 
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \cite{Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
Sitch-2003}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
of the most popular,
large-scale vegetation models lack any real demography or
size-structure \citep[see][]{Moorcroft-2001, Smith-2014}.

@@

At the same time, we expect the {\plant} model to be useful for
modelling other demographic phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size-structured
competition within a metapopulation of ``patches''. However, whereas
these models rely on simulations and were designed to capture large-scale regional
biogeochemistry driven by climatic variables,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied deterministically and in detail. In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. The package is thus designed to bridge the gap between 
a large-scale coupled simulation model and simple
abstract theoretical device.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. Being driven by traits, the model
can be extended to potentially very many species. Users can also add
their own physiological models, though this is a work in progress.

On top of this physiological model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
We model the emergent behaviour of patches deterministically, based on a
continual  size-density distribution. As such, it is not required to specify a
particular patch size.

@@

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is calculated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
The key outputs needed by higher levels (see below) are height growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011,Falster-2015}, but with extensions allowing for
diameter growth also to be estimated. To properly model trait
variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs.

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Table 3 in Appendix \ref{sec:FFW16}); however for some
parameters limited or no data were readily available. Providing
updated parameter sets is trivial (see  Appendix \ref{sec:parameters}); 
the challenge is to infer suitable
values by constraining outputs from {\plant} against diverse data
sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
are still actively working and one of the major use cases we envisage
for {\plant}.

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed and the higher level 
functionality of the package can still be used.  As an intermediate level, 
we allow for changes in key parameters to trigger changes 
in other parameters (i.e. hyper-parametrisation),
allowing straightforward modelling of trade-offs.  
For example, the trait \textsc{lma} is used to estimate the
rate of leaf turnover, based on an observed scaling relationship spanning
across diverse vegetation and plant functional types
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

@@

Ideally one would also consider spatial interactions within patches,
however such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980, 
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (\textsc{ebt})
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves the \textsc{pde}
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
\textsc{pde}. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{ebt}: i) a new method for estimating
the size-density distribution and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light (see Appendix \ref{sec:demography} for
details). The original \textsc{ebt} \citep{Deroos-1988,
  Deroos-1992, Deroos-1997} proceeds by approximating the first and
second moments of the density distribution \(n\) within each
cohort. In {\plant} we use a different approach. Instead of tracking
first and second moments of the size-distribution within cohorts, we
track a point mass estimate of \(n\) along the trajectory for the
boundary of each cohort. We found this approach preferable because it
does not require us to maintain a separate set of equations for the
smallest cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original \textsc{ebt}, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
problem later during patch development because much of the population
can be located between widely-spaced cohorts, delivering low numerical
precision. To maintain numerical precision we therefore use an
iterative algorithm to adaptively refine the cohort spacing until the
trajectories of adjacent cohorts remain adequately resolved throughout
development of the patch (see Appendix \ref{sec:demography} for
details). This refinement causes an uneven distribution of cohort
introduction times across patch age (Fig.  \ref{fig:patch}a), with
considerably tighter spacing of cohorts in younger patches.

@@

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age 
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors, yielding emergent non-linear outcomes
 (Fig. \ref{fig:patch}c).

@@

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
\citep{White-1979, Bormann-1979, Coomes-2007, Chambers-2013}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}b). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and linked via seed dispersal (Fig.
\ref{fig:schematic}b). We also assume that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second \textsc{pde}
(see Appendix \ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation 
 (Fig. \ref{fig:patch}c).

The main numerical challenge in stepping from single patch to
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate 
the demography of the metapopulation, this in turn produces an ``output'' 
number of seeds. The equilibrium occurs when a given
input seed
rain produces an equivalent output seed rain. Finding this equilibrium requires
finding the root of a multi-dimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix
\ref{sec:equilibrium}).

An attractive  feature of the metapopulation concept implemented here
is that integrates equilibrium and non-equilibrium approaches to
modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
Falster-2011}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

@@

The dynamic equilibrium captured by the trait-, size- and 
patch-structured metapopulation model potentially resolves several
challenges faced by unstructured models. First, by scaling up from
patch to metapopulation, we close the demographic feedback loop to
create a self-regulating system. Unlike some vegetation models, there
is no need to set the density of seeds or seedlings, these arise as
outputs from the model, based on the combined effects of the model's
physiological rules, species traits, competitive interactions and the
prevailing disturbance regime.

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
as  emergent phenomena. Moreover, the predictions from {\plant} 
are for a frequency distribution of patch states across the landscape 
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size- and patch-structured models provide unique opportunities for linking with 
satellite and forest survey data \citep{Moorcroft-2001, Purves-2008}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate invasion fitness -- 
the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Geritz-1998,Dieckmann-2007}.
While ideally calculated as the long-term per capita growth rate, 
a more convenient
indicator of invasion fitness in structured metapopulations is given by the logarithm of the
basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. 
\(R\) is simply the total seed output of the new
type, averaged across the metapopulation. So a new type can invade when
 \(\log R > 0\).

With fitness defined, it becomes possible to model trait-based adaptive dynamics
and community assembly \citep{Falster-2015}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits (Fig. \ref{fig:fitness}). The
reproductive success of individuals possessing a particular trait value
depends on both the traits and densities of other species in the current
resident community. Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution. Stale coexistence is possible when 
species with different traits can mutually invade \citep{Geritz-1998}
Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies and represents a prediction for the 
trait mixture that might might be favoured by natural selection 
\citep{Geritz-1998,Falster-2015}.

>>>>

\newcommand{\plant}{\texttt{plant}}
\usepackage{graphicx}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography are mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how  
  physiological rules and functional trade-offs interact with competition and 
  disturbance regimes to influence vegetation demography, structure and 
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and 
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. 
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
Ontogenetic 
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates 
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for vegetation dynamics and
diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
% TODO: Andrea notes that this section does not flow well.  I agree,
% but let's leave it until after submission.
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
Sitch-2003}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
of the most popular,
large-scale vegetation models lack any real demography or
size-structure \citep[see][]{Moorcroft-2001, Smith-2014}.

@@

At the same time, we expect the {\plant} model to be useful for
modelling other demographic phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size-structured
competition within a metapopulation of ``patches''. However, while
these models rely on simulations and were designed to capture large-scale regional
biogeochemistry driven by climatic variables, we designed
{\plant} so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied deterministically and in detail. In {\plant}, it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. The package is thus designed to bridge the gap between 
a large-scale coupled simulation model and a simple
abstract theoretical device.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. By being driven by traits, the
model can be extended to potentially very many species. Users will
also be able to add their own physiological models, though this is a
work in progress.

On top of this physiological model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
We model the emergent behaviour of patches deterministically, based on a
continual  size-density distribution. As such, specifying a patch size
is not required.

@@

% NOTE: Add something saying: we define a species as a group of
% individuals with identical traits?
The core of {\plant} is a model for an individual species'
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and the current light
environment. Gross photosynthetic income is calculated from the total
leaf area and the light distribution across the plant's canopy. Costs
of tissue respiration and turnover are then subtracted. The remaining
biomass is then allocated between growth and reproduction.  The key
outputs needed by higher levels of the model (see below) are height
growth rate, mortality rate and rate of seed production.  However
additional quantities are also computed, such as total assimilation
and the respiration, turnover and allocation rates for different
tissues (see Appendix \ref{sec:plant})

% TODO: Might need example/citation for trade-offs here?
The default physiological model used in {\plant} largely reflects that
presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. To properly model
trait variation, different parameters (such as rates of leaf turnover,
respiration, photosynthesis per leaf area) in the model are linked via
assumed trade-offs (see Appendix \ref{sec:parameters}).  We refer to
this default model as \texttt{FFW16}, reflecting our initials in this
paper.

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available
literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
some parameters were limited. Providing updated parameter sets is
trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
suitable values by constraining outputs from {\plant} against diverse
data sources \citep{Lebauer-2012, Keenan-2013}. This is an area in
which we are still actively working and one of the major use cases we
envisage for {\plant}.

While we have implemented one physiological model (\texttt{FFW16}),
the package is designed to allow arbitrary additional physiologies to
be introduced.  That is, rather than varying the parameters of the
equations, the equations themselves can be changed and the higher
level functionality of the package can still be used.  As an
intermediate level, we allow for changes in key parameters to trigger
changes in other parameters (i.e. hyper-parametrisation), allowing
straightforward modelling of trade-offs.  For example, the trait
\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
rate of leaf turnover, based on an observed scaling relationship
spanning across diverse vegetation and plant functional types
\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).

@@

Ideally, one would also consider spatial interactions within patches,
however such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980, 
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

@@

Our approach for modelling size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(n\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

We introduce two extensions to the \textsc{ebt}: i) a new method for
estimating the size-density distribution and ii) a technique for
handling strong size-asymmetric competitive feedbacks, such as occur
under strong competition for light (see Appendix \ref{sec:demography}
for details). The original \textsc{ebt} proceeds by approximating the
first and second moments of the density distribution \(n\) within each
cohort \citep{Deroos-1997}.
%
In {\plant} we use a different approach. Instead of tracking first and
second moments of the size-distribution within cohorts, we track a
point mass estimate of \(n\) along the trajectory for the boundary of
each cohort. We found this approach preferable because it does not
require us to maintain a separate set of equations for the smallest
cohort.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original \textsc{ebt}, cohorts were
introduced at a fixed time interval. Under competition, the growth
trajectories of individuals born at similar times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography}, \ref{sec:cohort-spacing} and
\ref{sec:patch}). This can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see Appendix
\ref{sec:demography} for details). This refinement causes an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with considerably tighter spacing of cohorts in
younger patches.

@@

Modelling of size-structured dynamics via solving of a deterministic
\textsc{pde} enables multiple demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch
development after a disturbance, including self-thinning and
successional replacement. Whereas previous approaches for modelling
self-thinning were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates
multiple species with different traits (Fig. \ref{fig:schematic}b;
Fig.  \ref{fig:patch}). Second, {\plant} allows for the effects of
competition and succession on productivity and biomass accumulation
within a patch to be investigated \citep{Falster-2011}. Leaf area
cover and rates of biomass production tend to vary with patch age
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. 
% TODO: I really don't know what the last bit of this sentence is
% meant to be saying.  Obviously it's something to do with the
% "emergent" properties of the patch, but not clear to me.
{\plant} enables the effects of competitive feedbacks to be
incorporated, in addition to any intrinsic physiological factors,
yielding emergent non-linear outcomes (Fig. \ref{fig:patch}c).

@@

% TODO: I removed the four(!) references here as I don't think this is
% controversial.
All vegetation is subject to a disturbance regime, such that patches
of the landscape are cleared at some rate, by fire, cyclone od
landslide.  The vegetation thus comprises a series of patches
differing in time since last disturbance and linked via seed dispersal
(Fig.  \ref{fig:schematic}b). Such a system is often referred to as a
structured metapopulation \citep{Gyllenberg-2001}.

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and sharing a common seed dispersal pool
(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
that the disturbances remove all established vegetation.  With this
approach we can scale up from simulating a single patch to simulate an
entire metapopulation.  The frequency-density \(p(a)\) of patches aged
\(a\) in the landscape is needed. Assuming there are a large number of
patches means the dynamics of \(p\) behave deterministically and can
be approximated via a second \textsc{pde} (see Appendix
\ref{sec:demography} for details). Scaling from patch to
metapopulation is then achieved by weighting the temporal dynamics of
a single patch by \(p(a)\), the relative abundance of patches aged
\(a\) in the metapopulation (Fig. \ref{fig:patch}c).

The main numerical challenge in stepping from a single patch to a
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An ``input'' seed rain is required to simulate the
demography of the metapopulation, this in turn produces an ``output''
number of seeds. The equilibrium occurs when a given input seed rain
produces an equivalent output seed rain. Finding this equilibrium
requires finding the root of a multi-dimensional function, subject to
the constraint that this root is stable (e.g., the trivial equilibrium
of zero seed rain always exists, but is often not stable; Appendix
\ref{sec:equilibrium}).

An attractive feature of the metapopulation concept implemented here
is that integrates both equilibrium and non-equilibrium approaches to
modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001,
  Falster-2011}. An equilibrium may be approached at the level of
metapopulation, meaning the seed rain and size structure across the
entire metapopulation is approximately stable. Yet the structure of
vegetation within individual patches is constantly in flux. Patches
continue to age and their residents continue to grow until they are
disturbed.

@@

The dynamic equilibrium captured by the trait-, size- and
patch-structured metapopulation model potentially resolves several
challenges faced by unstructured models. First, by scaling up from
patch to metapopulation, we close the demographic feedback loop to
create a self-regulating system. Unlike some vegetation models, there
is no need to set the density of seeds or seedlings.  These densities
arise as outputs from the model, based on the combined effects of the
model's physiological rules, species traits, competitive interactions
and the prevailing disturbance regime.

% NOTE: Andrea points out that young patches may not be the most
% common patch type in the metapopulation - it will depend on the
% regime.  And on the definition of young.
Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches tend to be more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
as  emergent phenomena. Moreover, the predictions from {\plant} 
are for a frequency distribution of patch states across the landscape 
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size- and patch-structured models provide unique opportunities for linking with 
satellite and forest survey data \citep{Moorcroft-2001, Purves-2008}.

Finally, with the extension to the metapopulation level, we are
modelling demography across the entire life cycle and are thus in a
position to estimate invasion fitness -- the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities
\citep{Geritz-1998,Dieckmann-2007}.  While ideally calculated as the
long-term per capita growth rate, a more convenient indicator of
fitness in structured metapopulations is given by the
logarithm of the basic reproduction ratio, \(R\)
\citep{Gyllenberg-2001, Metz-2001}.
% TODO: Ideally, don't start a sentence with an abbreviation
\(R\) is simply the total seed output of the new type, averaged across
the metapopulation. So a new type can invade when \(\log R > 0\).

With this measure of fitness, we can model trait-based adaptive
dynamics and community assembly \citep{Falster-2015}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits
(Fig. \ref{fig:fitness}). The reproductive success of individuals
possessing a particular trait value depends on both the traits and
densities of other species in the current resident community. Fitness
landscapes will be positive in regions that allow invasion, and zero
for species that are at equilibrium. The slope of the fitness
landscape with respect to traits indicates the direction of selection
(dashed line in Fig. \ref{fig:fitness}a) indicates selection for
smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution. Stable coexistence is
possible when species with different traits can mutually invade
\citep{Geritz-1998} Following the fitness gradient may lead to an
evolutionary branching point \citep{Geritz-1998,Dieckmann-2007}; at
this point there is no directional selection (solid line in
Fig. \ref{fig:fitness}a) but invasion is possible both above and below
it in trait space. It is often possible to find a configuration where
fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
except for the zero points where species are present. This community
is immune to invasion from any other strategies and represents a
prediction for the trait mixture that might might be favoured by
natural selection \citep{Geritz-1998,Falster-2015}.
